\propertitle

### 2019-06-16 maxNt_v5.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-16T12:46:55+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The context we consider is as follows. We have recorded the spiking
activities of $n$ neurons for a given amount of time. These spikes are
binned into $T$ time bins and binarized to $\set{0,1}$ values in each bin.
Denote by $\ya_{t}$ the number of neurons that fire during time bin $t$,
divided by the total number of neurons $n$; we call this the
\emph{normalized total activity} or just \enquote{total activity} for
short. Obviously $\ya_{t} \in\set{0,1/n,\dotsc,(n-1)/n,1}$; if $\ya_{t}=0$,
no neuron spiked during time bin $t$; if $\ya_{t}=1$, all spiked at some
point during time bin $t$. For brevity, let's say \enquote{at $t$} for
\enquote{during time bin $t$}.

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain regions
\mynote{be more specific}. The neurons that are recorded within the region
are not, and cannot yet be, specifically chosen from among the rest, owing
to several limiting factors -- for example, limitations in the precision
with which electrodes are inserted, or with which neurons are targeted by
viruses, and so on. In fact, the set of recorded neurons may even vary
during very long recordings or across experimental trials. We assume that
there's a small region, comprising a population of $N$ neurons, from which
other sets of neurons had the same probability of being recorded as the set
of $n$ neurons that have actually been recorded. During the recording, the
normalized total activity of these $N$ neurons at $t$ was $\yA_{t}$. We
don't know the values $\yA_{t}$; we only know for certain that
$\yA_{t} \in \set{0,1/N,\dotsc,(N-1)/N,1}$ and that $N\yA_{t} \ge n\ya_{t}$
for obvious reasons. For the moment we assume that we know $N$; we'll
discuss later the consequences of our lack of precise knowledge about $N$.

Our questions concern general features of the total activity $\yA$ of the
full population; for example: what was its time-frequency distribution? How
much activity should we expect in a time bin during the recording, or in a
new time bin? The answers to these questions are probabilistic and come in
the form of distributions.

The idea behind the approach presented here is easily summarized:
\begin{enumerate}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full population using the maximum-entropy method;
  \item\label{item:sample_step} the expectation constraints used in the
  method for the full population are, in turn, determined via sampling
  theory from the expectation constraints for the sample.
\end{enumerate}
Let's discuss these points in more detail.

Regarding point~\ref{item:maxent_step}, we assume you to be familiar with
the maximum-entropy method. In short, it amounts to a pair of
prescriptions: choose the distribution having maximum Shannon entropy among
those satisfying specific convex constraints, such as expectation values;
and judge these expectation values to be equal to measured averages. We add
two remarks that are seldom made in the literature. First, the
\enquote{distribution} given by this method is the zeroth-order
approximation of four different distributions:
\begin{itemize}
\item the \emph{probability} distribution for the value of the activity at any
  time bin among those \emph{recorded},
\item the \emph{probability} distribution for the value of the activity at
  a \emph{new} time bin,
\item the most probable \emph{time-frequency} distribution of the activity
  during the \emph{recording},
\item the most probable \emph{time-frequency} distribution of the activity
  in a very long run of \emph{new} time bins.
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
conceptually different inferences. Note that although the four
distributions above numerically coincide in this approximation, they
numerically differ in higher-order approximations. Second, the
maximum-entropy method based on the Shannon entropy implicitly makes
specific assumptions about the probability of the long-run frequency
distributions \citep{jaynes1986d_r1996,portamana2009,portamana2017}. We
discuss these assumptions in \sect***

In our case, to apply the method to the total activity of the full
population we need to fix some of its expectation values, for example its
moments. But we don't have any measured averages for the full population to
equate our expectations to. Here enters point~\ref{item:sample_step}: the
probability calculus gives an exact, linear relation between the first $m$
moments for the full population and the first $m$ for the sample; the ones
determine the others and vice versa at every time bin \citep[for
example][\eqns~(16)]{portamanaetal2015}. This is a classical result of
sampling theory \citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

@@

These two steps are even more straightforward if instead of power moments
we use \emph{normalized factorial moments}. The $m$th normalized factorial
moment is defined as the expectation
\begin{equation}
  \E\condt[\Bigg]{
    \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pf(\ya \| \yH),
\end{equation}
where $\ya\in\set{0, 1/n, 2/n, \dotsc, (n-1)/n, 1}$ is the normalized total
activity of a population of $n$ neurons, and $\pf(\ya \| \yH)$ is the
probability for activity $\ya$ given assumptions $\yH$. This moment can be
interpreted as the expectation of the number of distinct $m$-tuples of
simultaneously active neurons, normalized by the maximum possible number of
distinct $m$-tuples. For example, with $m=2$, if $na=4$ neurons are active
in a population of $n=5$, we have $\tbinom{4}{2}=6$ distinct pairs of
simultaneously active neurons, and if all five neurons were active we would
have a maximum of $\tbinom{5}{2}=10$ distinct pairs; the normalized number
of pairs is therefore $6/10$. Note that the first $m$ factorial moments
linearly determine the first $m$ moments and vice versa, so fixing the ones is
equivalent to fixing the others. Now assume that these $n$ neurons are a
sample from a full population of $N$ neurons, with possible normalized
activity $\yA \in\set{0,1/N,\dotsc,1}$. We have this extremely convenient
property: \emph{The first $n$ normalized factorial moments for the sample
  and for the full population are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad m \le n.
\end{equation}

We can hence apply the maximum-entropy method to obtain a distribution for
the full population by constraining its $m$th factorial moment to be equal
to the sample's recorded average of $\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$,
for as many $m$ as we please with $1\le m \le n$. If $\ya_{t}$ is the
normalized total activity of the sample during the time bin $t$, and the
recording with $T$ time bins, the constraint on the distribution
$\pf(\yA \| \yH)$ for the full population is
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA \| \yH)
\end{equation}
for any $m$s we wish.

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-06-16T21:39:12+0200
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Maximum-entropy distributions for a 
  neuronal population\\ from subpopulation data%\\{\large ***}%
}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Maximum-entropy from sample data}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

The context we consider is as follows. We have recorded the spiking
activities of $n$ neurons for a given amount of time. These spikes are
binned into $T$ time bins and binarized to $\set{0,1}$ values in each bin.
Denote by $\ya_{t}$ the number of neurons that fire during time bin $t$,
divided by the total number of neurons $n$; we call this the
\emph{normalized total activity} or just \enquote{total activity} or
\enquote{activity} for short. Obviously
$\ya_{t} \in\set{0,1/n,\dotsc,(n-1)/n,1}$; if $\ya_{t}=0$, no neuron spiked
during time bin $t$; if $\ya_{t}=1$, all spiked at some point during time
bin $t$. For brevity, let's say \enquote{at $t$} for \enquote{during time
  bin $t$}.

\mynote{maybe move this paragraph to the intro?}Present-day technologies
enable the recording of neuronal activity from small brain regions
\mynote{be more specific}. The neurons that are recorded within the region
are not, and cannot yet be, specifically chosen from among the rest, owing
to several limiting factors -- for example, limitations in the precision
with which electrodes are inserted, or with which neurons are targeted by
viruses, and so on. In fact, the set of recorded neurons may even change
during very long recordings or across experimental trials. We assume that
there's a region, comprising a population of $N$ neurons, from which other
sets of neurons had the same probability of being recorded as the set of
$n$ neurons that have actually been recorded. We call this the
\enquote{full population}. During the recording, the normalized total
activity of these $N$ neurons at $t$ was $\yA_{t}$. We don't know the
values $\yA_{t}$ at each $t$; we only know for certain that
$\yA_{t} \in \set{0,1/N,\dotsc,(N-1)/N,1}$ and that $N\yA_{t} \ge n\ya_{t}$
for obvious reasons. For the moment we assume that we know $N$; in \sect***
we discuss the consequences of our lack of precise knowledge about this
number.

Our questions concern general features of the total activity $\yA$ of the
full population during and after the recording. For example: what was its
frequency distribution in time? How much total activity should we expect at
a time bin during the recording, or at a new time bin? We cannot answer
these questions with certainty; we can only give a distribution of
probability over the possible answers. Our approach gives such a
probability distribution.

The idea behind the approach is easily summarized:
\begin{enumerate}[label=(\alph*)]
  \item\label{item:maxent_step} we build a distribution for the total
  activity of the full population using the maximum-entropy method;
  \item\label{item:sample_step} the expectation constraints used in the
  maximum-entropy method for the full population are, in turn, determined
  via sampling theory from the expectation constraints for the sample.
\end{enumerate}
Let's discuss these points in detail.

Regarding~\ref{item:maxent_step}, we assume you to be familiar with the
maximum-entropy method. In short, it amounts to a pair of prescriptions:
choose the distribution having maximum Shannon entropy among those
satisfying specific convex constraints, such as fixed expectation values; and
judge these expectation values to be equal to measured averages. We add two
remarks about this method that are seldom made in the literature. First,
the \enquote{distribution} given by this method is the zeroth-order
approximation of four different distributions:
\begin{itemize}
\item the \emph{probability} distribution for the value of the total
  activity at any time bin among those \emph{recorded},
\item the \emph{probability} distribution for the value of the total
  activity at a \emph{new} time bin,
\item the most probable \emph{time-frequency} distribution of the total
  activity across the \emph{recorded} bins,
\item the most probable \emph{time-frequency} distribution of the total
  activity in a very long run of \emph{new} time bins.
\end{itemize}
The maximum-entropy distribution is thus an approximate answer to four
conceptually different inferences. Note that although the four
distributions above numerically coincide in this approximation, they differ
in higher-order approximations. Second, the maximum-entropy method based on
the Shannon entropy implicitly makes specific assumptions about the
probability of the long-run frequency distributions
\citep{jaynes1986d_r1996,portamana2009,portamana2017}. We discuss these
assumptions in \sect***

In our case, to apply the method for the total activity of the full
population we need to fix some of the latter's expectation values, for
example its moments. But we don't have any measured averages for the full
population to equate our expectations to. Here enters
point~\ref{item:sample_step}: the probability calculus gives an exact,
linear relation between the first $m$ moments for the full population and
the first $m$ for the sample; the ones determine the others and vice versa
at every time bin \citep[for example][\eqns~(16)]{portamanaetal2015}. This
relation is a classical result of sampling theory
\citep[\chaps~I--IV]{whitworth1867_r1965}[\chap~II]{feller1950_r1968}[\chap~3]{jaynes1994_r2003}[see
also][]{whitworth1897}.

@@

These two steps are more straightforward if instead of power moments we use
\emph{normalized factorial moments}. The $m$th normalized factorial moment
of the probability distribution for the activity of the recorded neurons is
defined as the expectation
\begin{equation}
  \E\condt[\Bigg]{
    \binom{n \ya}{m}\,\binom{n}{m}^{-1}}{\yH}
  \equiv
  \sum_{\ya} %m!\,
  \binom{n \ya}{m}\,  \binom{n}{m}^{-1}\, \pf(\ya \| \yH),
  \qquad 1\le m \le n
\end{equation}
where $\pf(\ya \| \yH)$ is the probability for activity $\ya$ given
assumptions $\yH$; \enquote{$\ya$} may refer to any recorded bin or to a
new bin. This moment can be interpreted as the expectation of the number of
distinct $m$-tuples of simultaneously spiking neurons (within a bin's time
width), normalized by the number of distinct $m$-tuples. For example, with
$m=2$, if $na=4$ neurons spike in a population of $n=5$, we have
$\tbinom{4}{2}=6$ distinct pairs of simultaneously spiking neurons, and the
total number of distinct pairs is $\tbinom{5}{2}=10$. The normalized number
of pairs is therefore $6/10$. Note that the first $m$ factorial moments
linearly determine the first $m$ moments and vice versa, so fixing the ones
is equivalent to fixing the others. Now, since the recorded neurons are a
sample from the full population of $N$ neurons, we have this extremely
convenient property: \emph{The first $n$ normalized factorial moments for
  the sample and for the full population are numerically identical}:
\begin{equation}
  \label{eq:equal_m_factorialmoments}
  \E\condt[\big]{\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}}{\yH} =
  \E\condt[\big]{\tbinom{N\yA}{m}\;\tbinom{N}{m}^{-1}}{\yH},
  \qquad m \le n.
\end{equation}

We can therefore apply the maximum-entropy method to obtain a distribution
for the full population, by constraining its $m$th factorial moment to be
equal to the sample's recorded average of
$\tbinom{n\ya}{m}\;\tbinom{n}{m}^{-1}$, for as many $m$ as we please with
$1\le m \le n$. The constraints on the distribution $\pf(\yA \| \yH)$ for
the full population are
\begin{equation}
  \label{eq:final_constraint}
  \frac{1}{T}\sum_{t}\binom{n\ya_{t}}{m}\,\binom{n}{m}^{-1}
  =
   \sum_{\yA}
  \binom{N \yA}{m}\,  \binom{N}{m}^{-1}\, \pf(\yA \| \yH)
\end{equation}
for any $m$s we wish.

In the next section we apply this method to the data sets from two actual
recordings, and discuss the properties of the resulting distributions.

\section{Application: two data sets}
\label{sec:application}
